COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of introduction
Mention numerical integration in abstract
Copy editing of background

\begin{abstract}
The numerical solution of partial differential equations using the
finite element method is one of the key applications of high
performance computing. Local assembly is its characteristic
operation. This entails the execution of a problem-specific kernel for
each element in the discretized problem domain. Since the domain size
can be huge, executing efficient kernels is fundamental. Their
optimization is, however, a challenging issue. Even though affine loop
nests are generally present, the short trip counts and the complexity
of mathematical expressions make it hard to determine a single or
unique sequence of successful transformations. Therefore, we present
the design and systematic evaluation of COFFEE, a domain-specific
compiler for local assembly kernels. COFFEE manipulates abstract
syntax trees generated from a high-level domain-specific language for
PDEs by introducing domain-aware composable optimizations aimed at
improving instruction-level parallelism, especially SIMD
vectorization, and register locality. It then generates C code
including vector intrinsics. Experiments using a range of
finite-element forms of increasing complexity show that significant
performance improvement is achieved.
%
%The code is generated from a high level language for PDEs. An
%invocation of the kernel typically does tensor-like calculators over
%relatively small matrices; e.g., triply nested loops operating on
%about a dozen iterations. Because of the DSL context, COFFEE has a lot
%of flexibility on how it organizes/orders the calculations to tailor
%execution to a target architecture.
%
\end{abstract}

@@

Local assembly is the computation of contributions of a specific cell
in the discretized domain to the PDE solution. The process consists of
numerically evaluating problem-specific integrals to produce a matrix
and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose
sizes depend on the order of the method. This operation is applied to
all cells in the discretized domain. In this work we focus on local
matrices, or ``element matrices,'' which are more costly to compute
than element vectors.

Given a finite element description of the input problem, expressed
through the domain-specific Unified Form Language~\cite{ufl} (UFL),
Firedrake generates a C-code kernel implementing assembly using a
numerical integration algorithm. This kernel can be applied to any
element in the mesh: this descends from the mathematical formulation
of assembly, which states that the evaluation of a local matrix can be
reduced to integrating on a fixed ``reference'' element---a special
element that does not belong to the domain---after a suitable change
of coordinates. Firedrake triggers the compilation of an assembly
kernel using an available vendor compiler, and manages its (parallel)
execution over all elements in the mesh. As already explained, the
subject of this paper is to enhance this execution model by adding an
optimization stage prior to the generation of C code.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain, and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts: 1) calculation of the
Jacobian matrix, its determinant and its inverse: required for the
aforementioned change of coordinates, from the reference element to
the one being computed; 2) definition of basis functions: used,
intuitively, to interpolate the contribution to the PDE solution over
the element. The choice of the basis functions is expressed in UFL
directly by users. In the generated code, they are represented as
global read-only two dimensional arrays (i.e., using \texttt{static
  const} in C) of double precision floats; 3) evaluation of the
element matrix in an affine loop nest, in which the integration is
performed. Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, it has short trip counts (in the range 3--15), and the
computation reduces to a summation of a few products involving basis
functions. An example is provided in Listing~\ref{code:helmholtz},
which shows the assembly kernel for a Helmholtz problem using Lagrange
basis functions on 2D elements with polynomial order $p=1$. In other
scenarios, for instance when solving the Burgers equation, whose
assembly code is given in Listing~\ref{code:burgers}, the number of
arrays involved in the computation of the element matrix can be much
larger: in this case, 14 unique arrays are accessed, and the same
array can be referenced multiple times within the expression. This may
also require the evaluation of constants in outer loops (called $F$ in
the code) to act as scaling factors of arrays; trip counts can be
larger (proportionally to the order of the method); and arrays may be
block-sparse. In addition to a larger number of operations, more
complex cases like the Burgers equation are characterized by high
register pressure.

The Helmholtz and Burgers equations are exponents of a large class of
problems and, as such, will constitute our benchmark, along with the
Diffusion equation. We carefully motivate this choice in
Section~\ref{sec:real-examples}.

Despite the infinite variety of assembly kernels which Firedrake can
generate, it is still possible to identify common domain-specific
traits that can be exploited for effective code transformations and
SIMD vectorization. These include: 1) memory accesses along the three
loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k}
loops are interchangeable, whereas interchanges involving the $i$ loop
require pre-computation of values (e.g. the $F$ values in Burgers) and
introduction of temporary arrays, as explained in
Section~\ref{sec:code-transf}; 3) depending on the problem being
solved, the \texttt{j} and \texttt{k} loops could iterate over the
same iteration space; 4) most of the sub-expressions on the right hand
side of the element matrix computation depend on just two loops
(either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In
Section~\ref{sec:code-transf} we show how to exploit these
observations to define a set of systematic, composable optimizations.

>>>>

\begin{abstract}
The numerical solution of partial differential equations using the
finite element method is one of the key applications of high performance
computing. Local assembly is its characteristic operation. This entails
the execution of a problem-specific kernel to numerically evaluate an
integral for each element in the discretized problem domain. Since the
domain size can be huge, executing efficient kernels is fundamental.
Their optimization is, however, a challenging issue. Even though affine
loop nests are generally present, the short trip counts and the
complexity of mathematical expressions make it hard to determine a
single or unique sequence of successful transformations. Therefore, we
present the design and systematic evaluation of COFFEE, a
domain-specific compiler for local assembly kernels. COFFEE manipulates
abstract syntax trees generated from a high-level domain-specific
language for PDEs by introducing domain-aware composable optimizations
aimed at improving instruction-level parallelism, especially SIMD
vectorization, and register locality. It then generates C code including
vector intrinsics. Experiments using a range of finite-element forms of
increasing complexity show that significant performance improvement is
achieved.
%
%The code is generated from a high level language for PDEs. An
%invocation of the kernel typically does tensor-like calculators over
%relatively small matrices; e.g., triply nested loops operating on
%about a dozen iterations. Because of the DSL context, COFFEE has a lot
%of flexibility on how it organizes/orders the calculations to tailor
%execution to a target architecture.
%
\end{abstract}

@@

Local assembly is the computation of contributions of a specific cell in
the discretized domain to the linear system which yields the PDE
solution. The process consists of numerically evaluating
problem-specific integrals to produce a matrix and a vector
\cite{quadrature1,fluidity_manual_v4}, whose sizes depend on the order
of the method. This operation is applied to all cells in the discretized
domain. In this work we focus on local matrices, or ``element
matrices'', which are more costly to compute than element vectors.

Given a finite element description of the input problem, expressed
through the domain-specific Unified Form Language (UFL)~\cite{ufl},
Firedrake employs the FEniCS form compiler (FFC)~\cite{FFC-Compiler} to
generate a C-code kernel implementing assembly using a
numerical quadrature. This kernel can be applied to any
element in the mesh, which follows from a mathematical property of the
finite element method. The evaluation of a local matrix can be
reduced to integration on a fixed ``reference'' element --- a special
element that does not belong to the domain --- after a suitable change
of coordinates. Firedrake triggers the compilation of an assembly
kernel using an available vendor compiler, and manages its (parallel)
execution over all elements in the mesh. As already explained, the
subject of this paper is to enhance this execution model by adding an
optimization stage prior to the generation of C code.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain, and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts:
\begin{enumerate}
  \item Calculation of the Jacobian matrix, its determinant and its
    inverse required for the aforementioned change of coordinates from
    the reference element to the one being computed.
  % FIXME: this rather hand-wavy and not technically correct
  \item Definition of basis functions used, intuitively, to interpolate
    the contribution to the PDE solution over the element. The choice of
    basis functions is expressed in UFL directly by users. In the
    generated code, they are represented as global read-only two
    dimensional arrays (i.e., using \texttt{static const} in C) of
    double precision floats.
  \item Evaluation of the element matrix in an affine loop nest, in which
    the integration is performed.
\end{enumerate}
Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
unique arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionally to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

The Helmholtz and Burgers equations exemplify a large class of problems
and, as such, will constitute our benchmark problems, along with the
Diffusion equation. We carefully motivate this choice in
Section~\ref{sec:real-examples}.

Despite the infinite variety of assembly kernels which Firedrake can
generate, it is still possible to identify common domain-specific
traits that can be exploited for effective code transformations and
SIMD vectorization. These include: 1) memory accesses along the three
loop dimensions are always unit stride; 2) the \texttt{j} and \texttt{k}
loops are interchangeable, whereas interchanges involving the $i$ loop
require pre-computation of values (e.g. the $F$ values in Burgers) and
introduction of temporary arrays, as explained in
Section~\ref{sec:code-transf}; 3) depending on the problem being
solved, the \texttt{j} and \texttt{k} loops could iterate over the
same iteration space; 4) most of the sub-expressions on the right hand
side of the element matrix computation depend on just two loops
(either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In
Section~\ref{sec:code-transf} we show how to exploit these
observations to define a set of systematic, composable optimizations.
